<!DOCTYPE html>
<html lang="en"><head>
  <meta charset="utf-8">
  <meta http-equiv="X-UA-Compatible" content="IE=edge">
  <meta name="viewport" content="width=device-width, initial-scale=1"><!-- Begin Jekyll SEO tag v2.6.1 -->
<title>Feature selection part 1 | fastpages</title>
<meta name="generator" content="Jekyll v4.0.0" />
<meta property="og:title" content="Feature selection part 1" />
<meta property="og:locale" content="en_US" />
<meta name="description" content="A critique to univariate methods" />
<meta property="og:description" content="A critique to univariate methods" />
<link rel="canonical" href="https://david26694.github.io/david-masip-blog/experiments/lasso/lasso_vs_corr.Rmd" />
<meta property="og:url" content="https://david26694.github.io/david-masip-blog/experiments/lasso/lasso_vs_corr.Rmd" />
<meta property="og:site_name" content="fastpages" />
<script type="application/ld+json">
{"description":"A critique to univariate methods","@type":"WebPage","headline":"Feature selection part 1","url":"https://david26694.github.io/david-masip-blog/experiments/lasso/lasso_vs_corr.Rmd","@context":"https://schema.org"}</script>
<!-- End Jekyll SEO tag -->
<link rel="stylesheet" href="/david-masip-blog/assets/css/style.css"><link type="application/atom+xml" rel="alternate" href="https://david26694.github.io/david-masip-blog/feed.xml" title="fastpages" /><link rel="shortcut icon" type="image/x-icon" href="/david-masip-blog/images/favicon.ico"><!-- Begin Jekyll SEO tag v2.6.1 -->
<title>Feature selection part 1 | fastpages</title>
<meta name="generator" content="Jekyll v4.0.0" />
<meta property="og:title" content="Feature selection part 1" />
<meta property="og:locale" content="en_US" />
<meta name="description" content="A critique to univariate methods" />
<meta property="og:description" content="A critique to univariate methods" />
<link rel="canonical" href="https://david26694.github.io/david-masip-blog/experiments/lasso/lasso_vs_corr.Rmd" />
<meta property="og:url" content="https://david26694.github.io/david-masip-blog/experiments/lasso/lasso_vs_corr.Rmd" />
<meta property="og:site_name" content="fastpages" />
<script type="application/ld+json">
{"description":"A critique to univariate methods","@type":"WebPage","headline":"Feature selection part 1","url":"https://david26694.github.io/david-masip-blog/experiments/lasso/lasso_vs_corr.Rmd","@context":"https://schema.org"}</script>
<!-- End Jekyll SEO tag -->

<link href="https://unpkg.com/@primer/css/dist/primer.css" rel="stylesheet" />
<link rel="stylesheet" href="//use.fontawesome.com/releases/v5.0.7/css/all.css"><link type="application/atom+xml" rel="alternate" href="https://david26694.github.io/david-masip-blog/feed.xml" title="fastpages" />

<script src="https://cdnjs.cloudflare.com/ajax/libs/require.js/2.3.6/require.min.js"></script>
<script type="text/javascript">
require.config({
  paths: {
    jquery: 'https://code.jquery.com/jquery-3.5.0.min',
    plotly: 'https://cdn.plot.ly/plotly-latest.min'
  },

  shim: {
    plotly: {
      deps: ['jquery'],
      exports: 'plotly'
    }
  }
});
</script>

<script>
function wrap_img(fn) {
    if (document.attachEvent ? document.readyState === "complete" : document.readyState !== "loading") {
        var elements = document.querySelectorAll(".post img");
        Array.prototype.forEach.call(elements, function(el, i) {
            if (el.getAttribute("title") && (el.className != "emoji")) {
                const caption = document.createElement('figcaption');
                var node = document.createTextNode(el.getAttribute("title"));
                caption.appendChild(node);
                const wrapper = document.createElement('figure');
                wrapper.className = 'image';
                el.parentNode.insertBefore(wrapper, el);
                el.parentNode.removeChild(el);
                wrapper.appendChild(el);
                wrapper.appendChild(caption);
            }
        });
    } else { document.addEventListener('DOMContentLoaded', fn); }
}
window.onload = wrap_img;
</script>

<script>
    document.addEventListener("DOMContentLoaded", function(){
    // add link icon to anchor tags
    var elem = document.querySelectorAll(".anchor-link")
    elem.forEach(e => (e.innerHTML = '<i class="fas fa-link fa-xs"></i>'));
    });
</script>
</head><body><header class="site-header">

  <div class="wrapper"><a class="site-title" rel="author" href="/david-masip-blog/">fastpages</a><nav class="site-nav">
        <input type="checkbox" id="nav-trigger" class="nav-trigger" />
        <label for="nav-trigger">
          <span class="menu-icon">
            <svg viewBox="0 0 18 15" width="18px" height="15px">
              <path d="M18,1.484c0,0.82-0.665,1.484-1.484,1.484H1.484C0.665,2.969,0,2.304,0,1.484l0,0C0,0.665,0.665,0,1.484,0 h15.032C17.335,0,18,0.665,18,1.484L18,1.484z M18,7.516C18,8.335,17.335,9,16.516,9H1.484C0.665,9,0,8.335,0,7.516l0,0 c0-0.82,0.665-1.484,1.484-1.484h15.032C17.335,6.031,18,6.696,18,7.516L18,7.516z M18,13.516C18,14.335,17.335,15,16.516,15H1.484 C0.665,15,0,14.335,0,13.516l0,0c0-0.82,0.665-1.483,1.484-1.483h15.032C17.335,12.031,18,12.695,18,13.516L18,13.516z"/>
            </svg>
          </span>
        </label>

        <div class="trigger"><a class="page-link" href="/david-masip-blog/about/">About Me</a><a class="page-link" href="/david-masip-blog/experiments/lasso/lasso_vs_corr.Rmd">Feature selection part 1</a><a class="page-link" href="/david-masip-blog/experiments/lasso/lasso_vs_corr.html">Feature selection part 1</a><a class="page-link" href="/david-masip-blog/search/">Search</a><a class="page-link" href="/david-masip-blog/categories/">Tags</a></div>
      </nav></div>
</header>
<main class="page-content" aria-label="Content">
      <div class="wrapper">
        <article class="post h-entry" itemscope itemtype="http://schema.org/BlogPosting">

  <header class="post-header">
    <h1 class="post-title p-name" itemprop="name headline">Feature selection part 1</h1><p class="page-description">A critique to univariate methods</p><p class="post-meta post-meta-title"><time class="dt-published" datetime="" itemprop="datePublished">
        
      </time>
       • <span class="read-time" title="Estimated read time">
    
    
      3 min read
    
</span></p>

    
      <p class="category-tags"><i class="fas fa-tags category-tags-icon"></i></i> 
      
        <a class="category-tags-link" href="/david-masip-blog/categories/#markdown">markdown</a>
        
      
      </p>
    

    </header>

  <div class="post-content e-content" itemprop="articleBody">
    # Feature selection part 1: A critique to univariate methods

The purpose of this post is to show the weaknesses of univariate feature selection methods. This methods usually go in the following way:

- Compute a correlation metric for every feature with the target (this metric might be correlation or mutual information, among others)
- Select the features that score the best regarding this metric

A possible implementation is [SelectKBest in sklearn](https://scikit-learn.org/stable/modules/generated/sklearn.feature_selection.SelectKBest.html).

If you're an univariate method lover and have a different perspective, don't hesitate to contact me.

## Experiment set-up

Let's prepare an experiment to see the deficiencies of these methods. We generate two variables `x1` and `x2` and a target variable `y` that is `x1 + 0.1 * x2 + noise`. So, the variable `y` is very similar to `x1`, but has some `x2` as well. Then we create two variables `x1a` and `x2a`, that are noisy versions of `x1` and `x2`.


Although the data is artifical, I believe this occurs in real life. When performing feature engineering, we might end up with pretty good features that are correlated, and also have not so good features.

```{r message=FALSE, warning=FALSE, echo=FALSE}
# library(rmarkdown)
# render("experiments/lasso/lasso_vs_corr.Rmd", md_document(variant ='gfm', preserve_yaml=TRUE))
```
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(glmnet)
library(yardstick)

set.seed(42)

len <- 20000

x1 <- rnorm(len)
x2 <- rnorm(len)

# All the rnorm(len) that appear from now on are to simulate noise
y <- x1 + 0.1 * x2 + 0.05 * rnorm(len)

x1a <- 0.8 * x1 + 0.2 * rnorm(len)
x2a <- 0.8 * x2 + 0.2 * rnorm(len)

```


After this, we split the data in train and test sets to estimate errors. We also transform the data to matrix representation, as we're going to use _glmnet_ package that uses matrices to fit models.

```{r}
model_tbl <- tibble(
 y = y,
 x1 = x1,
 x2 = x2,
 x1a = x1a,
 x2a = x2a
)

# Split train and test ----------------------------------------------------

X <- as.matrix(select(model_tbl, -y))

X_train <- X[1:(len/2),]
y_train <- y[1:(len/2)]
X_test <- X[(len/2 + 1):len,]
y_test <- y[(len/2 + 1):len]

```


## Univariate-method approach

The univariate method approach that we follow here consists in keeping the most correlated variables with the target. For this reason, we compute the correlation for each variable:

```{r}
cor(X_train, y_train)
```

We'll keep only the first and third variables, as they are the most correlated with the target. We'll train a _ridge_ model using _glmnet_ and cross-validation to select the optimal hyperparameters:

```{r}
X_train_cor <- X_train[, c(1, 3)]
X_test_cor <- X_test[, c(1, 3)]

ridge <- cv.glmnet(X_train_cor, y_train, alpha = 0)

y_pred_cor <- as.vector(predict(ridge, X_test_cor, s = "lambda.min"))

```

## Multivariate methods: the Lasso

The Lasso can be thought of as a multivariate feature selection method, as it uses all the features at the same time to select a subset. Here we are applying a Lasso because `cv.glmnet` does Lasso (`alpha = 1`) by default.

```{r}
lasso <- cv.glmnet(X_train, y_train)

y_pred_lasso <- as.vector(predict(lasso, X_test, s = "lambda.min"))
```

## Results

### Coefficient comparison

By construction, the coefficients should be 1 for `x1` and 0.1 for `x2`. In the ridge case (recall that we've applied univariate selection using correlation), this won't happen as we've deleted `x2` because the correlation with the target was too low.

```{r}
coef(ridge, s = "lambda.min")
```

In fact, what we have is a mixture of `x1` and `x1a`.

On the other hand, the Lasso sets the coefficients of `x1a` and `x2a` to 0 (doesn't select those variables), and estimates the coefficents of `x1` and `x2` pretty well:

```{r}
coef(lasso, s = "lambda.min")
```

### Test set results

If you care more about model performance and not that much of proper estimation of the parameters, I also have an argument for you. The Lasso predictions are way better than the correlation method predictions:

```{r}
rmse_vec(y_pred_lasso, y_test)
rmse_vec(y_pred_cor, y_test)
```

We can even interpret it:

- The lasso error is around 0.05 which is the "noise size" used to generate `y`.
- The error of the univariate method is 0.15, which is 0.05 (noise size used to generate `y`) plus 0.10 (error we are commiting by not considering `x2`).


  </div><a class="u-url" href="/david-masip-blog/experiments/lasso/lasso_vs_corr.Rmd" hidden></a>
</article>
      </div>
    </main><footer class="site-footer h-card">
  <data class="u-url" href="/david-masip-blog/"></data>

  <div class="wrapper">

    <div class="footer-col-wrapper">
      <div class="footer-col">
        <p class="feed-subscribe">
          <a href="/david-masip-blog/feed.xml">
            <svg class="svg-icon orange">
              <use xlink:href="/david-masip-blog/assets/minima-social-icons.svg#rss"></use>
            </svg><span>Subscribe</span>
          </a>
        </p>
      </div>
      <div class="footer-col">
        <p>An easy to use blogging platform with support for Jupyter Notebooks.</p>
      </div>
    </div>

    <div class="social-links"><ul class="social-media-list"><li><a rel="me" href="https://github.com/fastai" title="fastai"><svg class="svg-icon grey"><use xlink:href="/david-masip-blog/assets/minima-social-icons.svg#github"></use></svg></a></li><li><a rel="me" href="https://twitter.com/fastdotai" title="fastdotai"><svg class="svg-icon grey"><use xlink:href="/david-masip-blog/assets/minima-social-icons.svg#twitter"></use></svg></a></li></ul>
</div>

  </div>

</footer>
</body>

</html>
